New cardinality estimation algorithms for HyperLogLog sketches

### 2017-02-02 paper/paper.tex
better algorithm for calculation of function tau with faster convergence
updated paper, some major changes

\title{New cardinality estimation algorithms for HyperLogLog sketches}
\author{Otmar Ertl \\ otmar.ertl@gmail.com}
\begin{document}
\maketitle
\let\thefootnote\relax\footnotetext{This paper together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}. A first version of this paper was published there on April 17, 2016.}
\begin{abstract}
This paper presents new methods to estimate the cardinalities of multisets recorded by HyperLogLog sketches. A theoretically motivated extension to the original estimator is presented that eliminates the bias for small and large cardinalities. Based on the maximum likelihood principle a second unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate. The maximum likelihood approach can also be applied to more than a single HyperLogLog sketch. In particular, it is shown that it gives more accurate cardinality estimates for the union, intersection, or relative complements of two sets that are both represented by HyperLogLog sketches compared to the conventional technique using the inclusion-exclusion principle. All the new methods are demonstrated and verified by extensive simulations.
\end{abstract}

@@

A more theoretically profound estimator for HyperLogLog sketches that does not depend on empirical data and significantly improves the estimation error is the historic inverse probability estimator \cite{Ting2014, Cohen2014}. It trades memory-efficiency for mergeability. The estimator needs to be continuously updated while inserting elements and the estimate depends on the element insertion order. Moreover, the estimator cannot be further used after merging two sketches, which limits its application to single data streams. If this restriction is okay, the self-learning bitmap \cite{Chen2011} 
could also be used, which provides a similar trade-off and is also 
more memory-efficient than the original HyperLogLog method.

Sometimes not only the number of distinct elements but also a sample of them is needed, in order to allow later filtering according to some predicate and estimating the cardinalities of corresponding subsets. In this case the k-minimum values algorithm \cite{Beyer2007} is the method of choice which also allows set manipulations like construction of intersections, disjoint sets, or unions \cite{Dasgupta2015}. The latter operation is the only one that is natively supported by HyperLogLog sketches. A sketch that represents the set operation result is not always needed. One approach to estimate the corresponding result cardinality directly is based on the inclusion-exclusion principle, which however can become quite inaccurate, especially for small Jaccard indices of the input sets \cite{Dasgupta2015}. Alternatively, the HyperLogLog sketches can be combined with some additional data structure that allows estimating the Jaccard index. Together with the HyperLogLog cardinality estimates the error of the intersection size estimate can be reduced \cite{Pascoe2013, Cohen2016}, however, at the expense of a significantly larger memory footprint due to the additional data structure.

\subsection{Outline}
This paper presents new algorithms for cardinality estimation from HyperLogLog sketches. In \cref{sec:hyperloglog_data_structure} we start with an introduction to HyperLogLog sketches and the corresponding update algorithm. We also discuss how the resulting state can be statistically described and approximated by a Poisson model.
In \cref{sec:cardinality_estimation} we describe the original cardinality estimation algorithm, its shortcomings at low and large cardinalities, and previous approaches to overcome them. We present a simple derivation of the original raw estimator and analyze the root cause for its limited operating range. We derive an improved version of the raw estimator, which leads to a new fast cardinality estimation algorithm that works over the full cardinality range as demonstrated by extensive simulations.
In \cref{sec:max_likelihood_estimation} we derive a second even more precise cardinality estimation algorithm based on the maximum likelihood principle which is again verified by simulations. As presented in \cref{sec:cardinality_estimation_set_intersections} the same approach can be generalized to two HyperLogLog sketches, which allows result cardinality estimation of set operations like intersections or complements. The simulation results show that the estimates are significantly better than those obtained by the inclusion-exclusion principle.
Finally, in \cref{sec:future_work} we discuss open problems and ideas for future work including HyperLogLog's potential use as locality-sensitve hashing algorithm before we conclude in \cref{sec:conclusion}.

\section{HyperLogLog data structure}
\label{sec:hyperloglog_data_structure}
The HyperLogLog algorithm uses a sketching data structure that consists of $\symNumReg$ registers. For performance reasons the number of registers $\symNumReg$ is chosen to be a power of 2, $\symNumReg = 2^\symPrecision$. $\symPrecision$ is the precision parameter that directly controls the relative error which scales like $1/\sqrt{\symNumReg}$. All registers start with zero initial value. Each element insertion potentially increases the value of one of these registers. The maximum value a register can reach is a natural bound given either by the output size of the used hash algorithm or the space that is reserved for a single register. Common implementations allocate up to 8 bits per register.

@@

\begin{algorithm}
\caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers. All registers $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ start from zero.}
\label{alg:insert}
\begin{algorithmic}
\Procedure {InsertElement}{\symDataItem}
\State $\langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision,\symBitRepB_1,\ldots,\symBitRepB_\symRegRange\rangle_2 \gets$ $(\symPrecision + \symRegRange)$-bit hash value of $\symDataItem$
\comm{0.25}{$\symBitRepA_\symIndexI,\symBitRepB_\symIndexI\in\{0,1\}$}
\State $\symRegVal \gets \min(\{\symS\mid \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
\comm{0.25}{$\symRegVal\in\lbrace 1,2,\ldots,\symRegRange+1\rbrace$}
\State $\symIndexI \gets 1+ \langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision\rangle_2$
\comm{0.25}{$\symIndexI\in\lbrace 1,2,\ldots,2^\symPrecision\rbrace$}
\If{$\symRegVal>\symRegValVariate_\symIndexI$}
\State $\symRegValVariate_\symIndexI \gets\symRegVal$
\EndIf
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

The order of register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ is not important for the estimation of the cardinality. More formally, the multiset $\lbrace\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg\rbrace$ is a sufficient statistic for $\symCardinality$.
Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg\rbrace = 0^{\symCountVariate_0}1^{\symCountVariate_1}\cdots\symRegRange^{\symCountVariate_{\symRegRange}}(\symRegRange+1)^{\symCountVariate_{\symRegRange+1}}$ where $\symCountVariate_\symRegVal$ is the multiplicity of value $\symRegVal$. As a consequence, the multiplicity vector $\boldsymbol{\symCountVariate} := (\symCountVariate_0,\ldots,\symCountVariate_{\symRegRange+1})$ is also a sufficient statistic for the cardinality. In addition, this vector contains all the information about the HyperLogLog sketch that is required for cardinality estimation. The two HyperLogLog parameters can be obtained by $\symPrecision = \log_2 (\sum_{\symRegVal=0}^{\symRegRange+1}\symCountVariate_\symRegVal
)$ and $\symRegRange=\dim\boldsymbol{\symCountVariate}-2$, respectively. As \cref{alg:sufficient_statistic} shows the calculation of the multiplicity vector $\boldsymbol{\symCountVariate}$ for a given HyperLogLog sketch with register values $\boldsymbol{\symRegValVariate}$ is straightforward.

@@

\subsection{Poisson approximation}
\label{sec:poisson_approximation}
Due to the statistical dependence of the register values, the probability mass function \eqref{equ:multinomialProbabilityMass} is difficult for further analysis. Therefore, a Poisson model can be used \cite{Flajolet2007}, which assumes that the cardinality itself is distributed according to a Poisson distribution
\begin{equation}
\symCardinality \sim \text{Poisson}(\symPoissonRate).
\end{equation}
Under the Poisson model the distribution of the register values is
\begin{align}
\symProbabilityMass(\boldsymbol{\symRegVal}\vert\symPoissonRate) 
&= 
\sum_{\symCardinality=0}^\infty \symProbabilityMass(\boldsymbol{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\label{equ:poissonization}
\\
&= 
\sum_{\symCardinality_1=0}^\infty
\cdots
\sum_{\symCardinality_\symNumReg=0}^\infty
\prod_{\symIndexI=1}^\symNumReg
\symRegProbability(\symRegVal_\symIndexI\vert\symCardinality_\symIndexI)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexI}}{\symCardinality_\symIndexI!\symNumReg^{\symCardinality_\symIndexI}}
\nonumber\\
&= 
\prod_{\symIndexI=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability(\symRegVal_\symIndexI\vert\symCardinality)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\nonumber\\
&= 
\prod_{\symRegVal=0}^{\symRegRange+1} \left(
\sum_{\symCardinality=0}^\infty
\symRegProbability(\symRegVal\vert\symCardinality)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\right)^{\!\!\symCount_\symRegVal}
\nonumber\\
\label{equ:poisson_pmf}
&=
e^{-\symCount_0\frac{\symPoissonRate}{\symNumReg}}
\left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\!\symCount_\symRegVal}\right)
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\!\symCount_{\symRegRange+1}}.
\end{align}
Here $\symCount_\symRegVal$ denotes the multiplicity of value $\symRegVal$  in the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$.
The final factorization shows that under the Poisson model the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ are independent and identically distributed. The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is defined by
\begin{equation}
\label{equ:register_value_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
=
\begin{cases}
0 & \symRegVal < 0 \\
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}} & 0\leq \symRegVal \leq \symRegRange \\
1 & \symRegVal > \symRegRange.
\end{cases}
\end{equation}

\subsection{Depoissonization}
\label{sec:depoissonization}
Due to the simpler probability mass function, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\boldsymbol{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ rather than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}. Depoissonization \cite{Jacquet1998} finally allows to translate the estimates back to the fixed-size model. Assume we have found an unbiased estimator for the Poisson rate
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = \symPoissonRate
\quad
\text{for all $\symPoissonRate\geq 0$}.
\end{equation}
We know from \eqref{equ:poissonization} 
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = 
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\end{equation}
and therefore
\begin{equation}
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
=
\symPoissonRate
\end{equation}
holds for all $\symPoissonRate\geq0$. The unique solution of this equation is given by
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symCardinality) = \symCardinality.
\end{equation}
Hence, the unbiased estimator $\symPoissonRateEstimate$ conditioned on $\symCardinality$ is also an unbiased estimator for $\symCardinality$, which motivates us to use $\symPoissonRateEstimate$ directly as estimator for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$. As simulation results will show later, the Poisson approximation works well over the entire cardinality range, even for estimators that are not exactly unbiased.

@@

\cref{fig:raw_estimate} shows the distribution of the relative error for the raw estimator as function of the cardinality. The chart is based on \num{10000} randomly generated HyperLogLog sketches. More details of the experimental setup will be explained later in \cref{sec:corrected_raw_estimation_error}. Obviously, the raw estimator is biased for small and large cardinalities where it fails to return accurate estimates. In order to cover the entire range of cardinalities, corrections for small and large cardinalities have been proposed.

@@

@@

@@

\left(1-\symX^{2^{-\symRegVal}}\right)
2^{-\symRegVal}.

We can cross-check the new estimator for the linear counting case with $\symRegRange=0$. Using the identity $\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = \symAlpha_\infty\symPowerSeriesFunc(\log_2(\log(1/\symX)))/\log(1/\symX)$, we get
\begin{equation}
\symPoissonRateEstimate 
= 
\frac{\symAlpha_\infty\symNumReg}{
\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)
+
\symLargeCorrectionFunc(\symCountVariate_0/\symNumReg)
}
=
\frac{\symNumReg\log\!\left(\symNumReg/\symCountVariate_0\right)}{\symPowerSeriesFunc\!\left(\log_2\!\left(\log\!\left(\symNumReg/\symCountVariate_0\right)\right)\right)}
\end{equation}
which is as expected almost identical to the linear counting estimator \eqref{equ:linear_counting_estimator}, because $\symPowerSeriesFunc(\symX)\approx 1$ (see \cref{sec:derivation_raw_estimator}).
\subsection{Improved raw estimation algorithm}
The improved raw estimator \eqref{equ:correctedestimator} leads to a new cardinality estimation algorithm for HyperLogLog sketches. \cref{alg:corrected_raw_estimation} demonstrates the numerical evaluation of the estimator. The values of functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ can be either calculated on-demand or pre-calculated. Possible values of $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$ are $\{0,1,\ldots, \symNumReg\}$. Therefore, if performance matters, the function values can be precalculated for all possible arguments and kept in lookup tables of size $\symNumReg+1$.

>>>>

\title{New cardinality estimation algorithms for HyperLogLog sketches}
\author{Otmar Ertl \\ otmar.ertl@gmail.com}
\begin{document}
\maketitle
\let\thefootnote\relax\footnotetext{This paper together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}. A first version of this paper was published there on April 17, 2016.}
\begin{abstract}
This paper presents new methods to estimate the cardinalities of multisets recorded by HyperLogLog sketches. A theoretically motivated extension to the original estimator is presented that eliminates the bias for small and large cardinalities. Based on the maximum likelihood principle a second unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate. The maximum likelihood approach can also be applied to more than a single HyperLogLog sketch. In particular, it is shown that it gives more precise cardinality estimates for union, intersection, or relative complements of two sets that are both represented by HyperLogLog sketches compared to the conventional technique using the inclusion-exclusion principle. All the new methods are demonstrated and verified by extensive simulations.
\end{abstract}

@@

A more theoretically profound estimator for HyperLogLog sketches that does not depend on empirical data and significantly improves the estimation error is the historic inverse probability estimator \cite{Ting2014, Cohen2014}. It trades memory-efficiency for mergeability. The estimator needs to be continuously updated while inserting elements and the estimate depends on the element insertion order. Moreover, the estimator cannot be further used after merging two sketches, which limits its application to single data streams. If this restriction is acceptable, the self-learning bitmap \cite{Chen2011} 
could also be used, which provides a similar trade-off and also needs less space than the original HyperLogLog method to achieve the same precision.

Sometimes not only the number of distinct elements but also a sample of them is needed in order to allow later filtering according to some predicate and estimating the cardinalities of corresponding subsets. In this case the k-minimum values algorithm \cite{Beyer2007} is the method of choice which also allows set manipulations like the construction of intersections, disjoint sets, or unions \cite{Dasgupta2015}. The latter operation is the only one that is natively supported by HyperLogLog sketches. A sketch that represents the set operation result is not always needed. One approach to estimate the corresponding result cardinality directly is based on the inclusion-exclusion principle, which however can become quite inaccurate, especially for small Jaccard indices of the input sets \cite{Dasgupta2015}. Alternatively, the HyperLogLog sketches can be combined with some additional data structure that allows estimating the Jaccard index. Together with the HyperLogLog cardinality estimates the error of the intersection size estimate can be reduced \cite{Pascoe2013, Cohen2016}, however, at the expense of a significantly larger memory footprint due to the additional data structure.

\subsection{Outline}
This paper presents new algorithms for cardinality estimation from HyperLogLog sketches. In \cref{sec:hyperloglog_data_structure} we start with an introduction to HyperLogLog sketches and the corresponding update algorithm. We also discuss how the resulting state can be statistically described and approximated by a Poisson model.
In \cref{sec:cardinality_estimation} we describe the original cardinality estimation algorithm, its shortcomings at low and large cardinalities, and previous approaches to overcome them. We present a simple derivation of the original raw estimator and analyze the root cause for its limited operating range. We derive an improved version of the raw estimator, which leads to a new fast cardinality estimation algorithm that works over the full cardinality range as demonstrated by extensive simulations.
In \cref{sec:max_likelihood_estimation} we derive a second even more precise cardinality estimation algorithm based on the maximum likelihood principle which is again verified by simulations. As presented in \cref{sec:cardinality_estimation_set_intersections} the same approach can be generalized to two HyperLogLog sketches, which allows result cardinality estimation of set operations like intersections or complements. The simulation results show that the estimates are significantly better than those obtained using the inclusion-exclusion principle.
Finally, in \cref{sec:future_work} we discuss open problems and ideas for future work including HyperLogLog's potential application as locality-sensitve hashing algorithm before we conclude in \cref{sec:conclusion}.

\section{HyperLogLog data structure}
\label{sec:hyperloglog_data_structure}
The HyperLogLog algorithm uses a sketching data structure that consists of $\symNumReg$ registers. For performance reasons the number of registers $\symNumReg$ is chosen to be a power of 2, $\symNumReg = 2^\symPrecision$. $\symPrecision$ is the precision parameter that directly controls the relative error which scales like $1/\sqrt{\symNumReg}$ \cite{Flajolet2007}. All registers start with zero initial value. Each element insertion potentially increases the value of one of these registers. The maximum value a register can reach is a natural bound given either by the output size of the used hash algorithm or the space that is reserved for a single register. Common implementations allocate up to 8 bits per register.

@@

\begin{algorithm}
\caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers. All registers $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ start from zero. $\langle\ldots\rangle_2$ denotes the binary representation of an integer.}
\label{alg:insert}
\begin{algorithmic}
\Procedure {InsertElement}{\symDataItem}
\State $\langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision,\symBitRepB_1,\ldots,\symBitRepB_\symRegRange\rangle_2 \gets$ $(\symPrecision + \symRegRange)$-bit hash value of $\symDataItem$
\comm{0.25}{$\symBitRepA_\symIndexI,\symBitRepB_\symIndexI\in\{0,1\}$}
\State $\symRegVal \gets \min(\{\symS\mid \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
\comm{0.25}{$\symRegVal\in\lbrace 1,2,\ldots,\symRegRange+1\rbrace$}
\State $\symIndexI \gets 1+ \langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision\rangle_2$
\comm{0.25}{$\symIndexI\in\lbrace 1,2,\ldots,\symNumReg\rbrace$}
\If{$\symRegVal>\symRegValVariate_\symIndexI$}
\State $\symRegValVariate_\symIndexI \gets\symRegVal$
\EndIf
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

The order of register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ is not important for the estimation of the cardinality. More formally, the multiset $\lbrace\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg\rbrace$ is a sufficient statistic for $\symCardinality$.
Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg\rbrace = 0^{\symCountVariate_0}1^{\symCountVariate_1}\cdots\symRegRange^{\symCountVariate_{\symRegRange}}(\symRegRange+1)^{\symCountVariate_{\symRegRange+1}}$ where $\symCountVariate_\symRegVal$ is the multiplicity of value $\symRegVal$. As a consequence, the multiplicity vector $\boldsymbol{\symCountVariate} := (\symCountVariate_0,\ldots,\symCountVariate_{\symRegRange+1})$ is also a sufficient statistic for the cardinality. In addition, this vector contains all the information about the HyperLogLog sketch that is required for cardinality estimation. The two HyperLogLog parameters can be obtained by $\symPrecision = \log_2 (\sum_{\symRegVal=0}^{\symRegRange+1}\symCountVariate_\symRegVal
)$ and $\symRegRange=\dim\boldsymbol{\symCountVariate}-2$, respectively. \cref{alg:sufficient_statistic} shows the calculation of the multiplicity vector $\boldsymbol{\symCountVariate}$ for a given HyperLogLog sketch with register values $\boldsymbol{\symRegValVariate}$.

@@

\subsection{Poisson approximation}
\label{sec:poisson_approximation}
Due to the statistical dependence of the register values, the probability mass function \eqref{equ:multinomialProbabilityMass} makes further analysis difficult. Therefore, a Poisson model can be used \cite{Flajolet2007}, which assumes that the cardinality itself is distributed according to a Poisson distribution
\begin{equation}
\symCardinality \sim \text{Poisson}(\symPoissonRate).
\end{equation}
Under the Poisson model the distribution of the register values is
\begin{align}
\symProbabilityMass(\boldsymbol{\symRegVal}\vert\symPoissonRate) 
&= 
\sum_{\symCardinality=0}^\infty \symProbabilityMass(\boldsymbol{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\label{equ:poissonization}
\\
&= 
\sum_{\symCardinality_1=0}^\infty
\cdots
\sum_{\symCardinality_\symNumReg=0}^\infty
\prod_{\symIndexI=1}^\symNumReg
\symRegProbability(\symRegVal_\symIndexI\vert\symCardinality_\symIndexI)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexI}}{\symCardinality_\symIndexI!\symNumReg^{\symCardinality_\symIndexI}}
\nonumber\\
&= 
\prod_{\symIndexI=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability(\symRegVal_\symIndexI\vert\symCardinality)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\nonumber\\
&= 
\prod_{\symRegVal=0}^{\symRegRange+1} \left(
\sum_{\symCardinality=0}^\infty
\symRegProbability(\symRegVal\vert\symCardinality)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\right)^{\!\!\symCount_\symRegVal}
\nonumber\\
\label{equ:poisson_pmf}
&=
e^{-\symCount_0\frac{\symPoissonRate}{\symNumReg}}
\left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\!\symCount_\symRegVal}\right)
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\!\symCount_{\symRegRange+1}}.
\end{align}
Here $\symCount_\symRegVal$ denotes the multiplicity of value $\symRegVal$  in the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$.
The final factorization shows that under the Poisson model the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ are independent and identically distributed. The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is given by
\begin{equation}
\label{equ:register_value_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
=
\begin{cases}
0 & \symRegVal < 0 \\
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}} & 0\leq \symRegVal \leq \symRegRange \\
1 & \symRegVal > \symRegRange.
\end{cases}
\end{equation}

\subsection{Depoissonization}
\label{sec:depoissonization}
Due to the simpler probability mass function, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\boldsymbol{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ than for the cardinality $\symCardinality$ under the fixed-size model \eqref{equ:multinomialProbabilityMass}. Depoissonization \cite{Jacquet1998} finally allows to translate the estimates back to the fixed-size model. Assume we have found an unbiased estimator for the Poisson rate
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = \symPoissonRate
\quad
\text{for all $\symPoissonRate\geq 0$}.
\end{equation}
We know from \eqref{equ:poissonization} 
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = 
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\end{equation}
and therefore
\begin{equation}
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
=
\symPoissonRate
\end{equation}
holds for all $\symPoissonRate\geq0$. The unique solution of this equation is given by
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symCardinality) = \symCardinality.
\end{equation}
Hence, the unbiased estimator $\symPoissonRateEstimate$ conditioned on $\symCardinality$ is also an unbiased estimator for $\symCardinality$, which motivates us to use $\symPoissonRateEstimate$ directly as estimator for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$. As simulation results will show later, the Poisson approximation works well over the entire cardinality range, even for estimators that are not exactly unbiased.

@@

\cref{fig:raw_estimate} shows the distribution of the relative error for the raw estimator as function of the cardinality. The chart is based on \num{10000} randomly generated HyperLogLog sketches. More details of the experimental setup will be explained later in \cref{sec:experimental_setup}. Obviously, the raw estimator is biased for small and large cardinalities where it fails to return accurate estimates. In order to cover the entire cardinality range, corrections for small and large cardinalities have been proposed.

@@

\label{equ:tau}

@@

%\begin{equation}
%\symLargeCorrectionFunc(\symX)
%:=
%\sum_{\symRegVal=1}^\infty
%\symX^{2^{-\symRegVal}}
%\left(1-\symX^{2^{-\symRegVal}}\right)
%2^{-\symRegVal}.
%\end{equation}
\label{equ:sigma}

@@

2^{-\symRegVal}
.
%=
%\frac{1}{3}
%\left(
%1
%-
%\symX
%-
%\sum_{\symRegVal=1}^\infty
%\left(
%1-
%\symX^{2^{-\symRegVal}}
%\right)^2
%2^{-\symRegVal}
%\right)

We can cross-check the new estimator for the linear counting case with $\symRegRange=0$. Using the identity 
\begin{equation}
\label{equ:sigma_tau_relationship}
\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = \symAlpha_\infty\symPowerSeriesFunc(\log_2(\log(1/\symX)))/\log(1/\symX),
\end{equation}
 we get
\begin{equation}
\symPoissonRateEstimate 
= 
\frac{\symAlpha_\infty\symNumReg}{
\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)
+
\symLargeCorrectionFunc(\symCountVariate_0/\symNumReg)
}
=
\frac{\symNumReg\log\!\left(\symNumReg/\symCountVariate_0\right)}{\symPowerSeriesFunc\!\left(\log_2\!\left(\log\!\left(\symNumReg/\symCountVariate_0\right)\right)\right)}
\end{equation}
which is as expected almost identical to the linear counting estimator \eqref{equ:linear_counting_estimator}, because $\symPowerSeriesFunc(\symX)\approx 1$ (see \cref{sec:derivation_raw_estimator}).
\subsection{Improved raw estimation algorithm}
The improved raw estimator \eqref{equ:correctedestimator} can be directly translated into a new cardinality estimation algorithm for HyperLogLog sketches as shown in \cref{alg:corrected_raw_estimation}. Since the series \eqref{equ:sigma} converges quadratically for all $\symX\in[0,1)$ and its terms can be recursively calculated using elementary operations, the function $\symSmallCorrectionFunc$ can be quickly calculated to machine precision. The case $\symX=1$ must be handled separately, because the series diverges and causes an infinite denominator in \eqref{equ:correctedestimator} and therefore a vanishing cardinality estimate. As this case only occurs if all register values are zero ($\symCountVariate_0=\symNumReg$), this is exactly what is expected. Among the remaining possible arguments $\lbrace0,\frac{1}{\symNumReg},\frac{2}{\symNumReg},\ldots,\frac{\symNumReg-1}{\symNumReg}\rbrace$ we have slowest convergence for $\symX=\frac{\symNumReg-1}{\symNumReg}$. However, even in this case we have a manageable amount of iteration cycles. For example, if double-precision floating-point arithmetic is used, the routine for calculating  $\symSmallCorrectionFunc$ converges after \num{18} and \num{26} iteration cycles for $\symPrecision=12$ and $\symPrecision=20$, respectively.

%
%The case $\symX=1$, for which the series does not converge, occurs only if all register values are zero ($\symCountVariate_0=\symNumReg$). However, the resulting positive infinite denominator in \eqref{equ:correctedestimator} yields zero for the cardinality estimate as expected. In order to estimate the number of iteration cycles needed to determine $\symSmallCorrectionFunc$ up to some precision $\symStopDelta$, we consider the case $\symX = 1-\frac{1}{\symNumReg}$ for which we have slowest convergence speed among all relevant argument values $\lbrace0,\frac{1}{\symNumReg},\frac{2}{\symNumReg},\ldots,1-\frac{1}{\symNumReg}\rbrace$. We can stop summation, if two subsequent terms differ by a factor more than $\symStopDelta$. This is the case, if $\frac{\symX^{2^{\symRegVal+1}} 2^{\symRegVal}}{\symX^{2^\symRegVal} 2^{\symRegVal-1}} < \symStopDelta$ which is equivalent to $
%\symRegVal > \frac{
%\log(\log(2/\symStopDelta))
%-
%\log(-\log(1-\frac{1}{\symNumReg}))
%}
%{
%\log(2)
%}
%$. Due to the inequality $-\log(1-\frac{1}{\symNumReg})\geq \frac{1}{\symNumReg}$ this is also satisfied, if $
%\symRegVal > \frac{
%\log(\log(2/\symStopDelta))
%}
%{
%\log(2)
%}
%+
%\symPrecision
%$
%holds. Therefore, not more than $\symPrecision + 5$ iterations are necessary to achieve double-precision $\symStopDelta = 2^{-53}$.
